---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Background

I love hillshade so much. It adds so much dimension to a map. I typically play around with hillshade in ArcGIS Pro but R packages have come a long way recently so I wanted to try out some techniques for hillshading inside R. 

I Followed this tutorial: <https://dominicroye.github.io/en/2022/hillshade-effects/> And this one: <https://dieghernan.github.io/202210_tidyterra-hillshade/>

# Loack Packages
```{r}
#Packages
if(!require("dplyr")) install.packages("dplyr") #
if(!require("ggplot2")) install.packages("ggplot2") #
if(!require("raster")) install.packages("raster")#
if(!require("elevatr")) install.packages("elevatr")#
if(!require("terra")) install.packages("terra")#
if(!require("tidyterra")) install.packages("tidyterra")#
if(!require("ggnewscale")) install.packages("ggnewscale") #
if(!require("here")) install.packages("here")
if(!require("tigris")) install.packages("tigris")#

# packages
library(sf)
library(raster)
library(elevatr)
library(dplyr)
library(ggplot2)
library(terra)
library(tidyterra)
library(ggnewscale)
library(tigris)
```

# Download state shapes
```{r}
# Get states geography. Filter out pacific islands
states <- tigris::states(cb = TRUE) # cb = catographic boundary
```

```{r}
# filter for your state
state <- states |>
  dplyr::filter(NAME =="North Carolina")

plot(state)
```

# Download Elevation Data

Use the shape of the state we just filtered as the location. The z-value is the zoom level and a 7 works well for US States.

```{r}
#Download elevation data using elevatr package
elevation <- elevatr::get_elev_raster(locations = state, z = 7, clip = "bbox") |>
  # And convert to terra format
  terra::rast() |>
  # Mask to the shape so the shape is exactly that of Nevada
  raster::mask(terra::vect(state)) |>
  #EPSG:6347 - NAD83 / UTM zone 18N
  terra::project("EPSG:4326")
  

# Plot
terra::plot(elevation)
```

```{r}
# convert the raster into a data.frame of xyz
elevation_df <- as.data.frame(elevation, xy = TRUE)

# Rename z to "alt"
names(elevation_df)[3] <- "alt"

# Turn negative values to 0
elevation_df <- elevation_df %>%
  mutate(alt = pmax(0, alt))

# first version of a map
ggplot() +
  geom_raster(data = elevation_df,
              aes(x, y, fill = alt)) +
  tidyterra::scale_fill_hypso_tint_c(breaks = seq(0, 2000, 400)) +
  guides(fill = guide_colorsteps(barwidth = 10,
                                 barheight = .5,
                                 title.position = "right")) +
  labs(fill = "m") +
  coord_sf() +
  theme_void() +
  theme(legend.position = "bottom")
```

# Build Hillshade
## Slope
```{r}
# estimate the slope
slope <- terra::terrain(elevation, "slope", unit = "radians")
plot(slope)
```

## Aspect
```{r}
# estimate the aspect or orientation
aspect <- terra::terrain(elevation, "aspect", unit = "radians")
plot(aspect)
```

## Hillshade
```{r}
# calculate the hillshade effect with 45ยบ of elevation
hillshade <- terra::shade(slope, aspect, 
      angle = 45, 
      direction = 300,
      normalize= TRUE)

# Palette
pal_greys <- hcl.colors(1000, "Grays")

#raw hillshade map 
ggplot() +
  geom_spatraster(data = hillshade, show.legend = FALSE) +
  scale_fill_gradientn(colors = pal_greys, na.value = NA) +
  theme_void() 
```

## Figure out elevation limits to guide plotting in ggplot
```{r}

# Min/Max limits of elevation
elev_limits <- minmax(elevation) %>% as.vector()

# Rounded to lower and upper 500
elev_limits <- c(floor(elev_limits[1] / 500), ceiling(elev_limits[2] / 500)) * 500

# And making min value to 0.
elev_limits <- pmax(elev_limits, 0)

# Compare
minmax(elevation) %>% as.vector()
elev_limits
# 0 - 2000

# Test plot with limits/ Try different palette colors from scale_fill_hypso_tint(). Palette colors are available here: https://dieghernan.github.io/tidyterra/reference/scale_hypso.html
ggplot() +
  geom_spatraster(data = elevation) +
  scale_fill_hypso_tint_c(
      limits = elev_limits,
      palette = "spain"
    ) +
    theme_minimal()
```

```{r}
ggplot() +# Add hillshade raster
  geom_spatraster(data = hillshade,  show.legend =  FALSE) +
  # Gray color scale for hillshade
  scale_fill_gradientn(colors = pal_greys, na.value = NA) +
  # Adds a new scale to ggplot so we can layer another raster
  ggnewscale::new_scale_fill() +
  # Add elevation raster
  geom_spatraster(data = elevation, maxcell = Inf) +
  scale_fill_hypso_tint_c(
    limits = elev_limits,
    palette = "utah_1",
    alpha = 0.4,
    # For the legend I use custom breaks
    breaks = c(seq(0, 500, 100), seq(750, 1500, 250), 2000)) +
  # Change legend guide
  guides(fill = guide_legend(
    title = "   m",
    direction = "horizontal",
    nrow = 1,
    keywidth = 1.75,
    keyheight = 0.5,
    label.position = "bottom",
    title.position = "right",
    override.aes = list(alpha = 1)
  )) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.spacing.x = unit(0, "pt"),
    # Transparent background for saving .png
    panel.background = element_rect(fill='transparent', color = NA),
    plot.background = element_rect(fill='transparent', color=NA),
    legend.background = element_rect(fill='transparent', color = NA),
    legend.box.background = element_rect(fill='transparent', color = NA)
       )
```

# Save a .png
```{r}
ggsave(filename = here::here("Images", "NC_hillshade.png"),  width = 8, height = 6, dpi = 600, device = "png")
```
